---
title: "biocMultiAssayToyExample"
author: "Marcel Ramos, Levi Waldron"
date: "September 4, 2015"
output: html_document
---

In this example we have 4 patients, and a bit of metadata on them:
```{r}
masPheno <- data.frame(sex=c("M", "F", "M", "F"),
						  age=38:41,
						  row.names=c("Jack", "Jill", "Bob", "Barbara"))
masPheno
```

We have three matrix-like datasets.  First let's say expression data:

```{r, message=FALSE}
library(affy)
(arraydat <- matrix(seq(101, 108), ncol=4, dimnames=list(c("ENST00000294241", "ENST00000355076"), c("array1", "array2", "array3", "array4"))))
arraypdat <- as(data.frame(slope53=rnorm(4), row.names=c("array1", "array2", "array3", "array4")), "AnnotatedDataFrame")
exprdat <- ExpressionSet(assayData=arraydat, phenoData=arraypdat)
exprdat
```

The following map matches masterPheno sample names to exprdata sample names.  Note that row orders aren't initially matched up.
```{r}
(exprmap <- data.frame(master=rownames(masPheno)[c(1, 2, 4, 3)], assay=c("array1", "array2", "array3", "array4")))
```

Now methylation data.  It uses gene identifiers also, but measures a partially overlapping set of genes.  For fun, let's store this as a simple matrix. Also, it contains a replicate for one of the patients.

```{r}
(methyldat <- matrix(1:10, ncol=5, 
                     dimnames=list(c("ENST00000355076", "ENST00000383706"),
                                   c("methyl1", "methyl2", "methyl3", "methyl4", "methyl5"))))
```

The following map matches masterPheno sample names to methyldat sample names.

```{r}
(methylmap <- data.frame(master = c("Jack", "Jack", "Jill", "Barbara", "Bob"),
                        assay = c("methyl1", "methyl2", "methyl3", "methyl4", "methyl5")))
```

Now we have a microRNA platform, which has no common identifiers.  It is also missing data for Jill.  Just for fun, let's use the same sample naming convention as we did for arrays.

```{r}
(microdat <- matrix(201:212, ncol=3, 
                    dimnames=list(c("hsa-miR-21", "hsa-miR-191", "hsa-miR-148a", "hsa-miR148b"), 
                                  c("array1", "array2", "array3"))))
```

And the following map matches masterPheno sample names to microdat sample names.
```{r}
(micromap <- data.frame(master = c("Jack", "Barbara", "Bob"),
                        assay = c("micro1", "micro2", "micro3")))
```

Let's include a GRangesList:  

```{r}
library(GenomicRanges)
gr1 <-
  GRanges(seqnames = "chr3", ranges = IRanges(58000000, 59502360), #completely encompasses ENST00000355076
          strand = "+", score = 5L, GC = 0.45)
gr2 <-
  GRanges(seqnames = c("chr3", "chr3"),
          ranges = IRanges(c(58493000, 3), width=9000), #first is within ENST0000035076
          strand = c("+", "-"), score = 3:4, GC = c(0.3, 0.5))
gr3 <-
  GRanges(seqnames = c("chr1", "chr2"),
          ranges = IRanges(c(1, 4), c(3, 9)),
          strand = c("-", "-"), score = c(6L, 2L), GC = c(0.4, 0.1))
grl <- GRangesList("gr1" = gr1, "gr2" = gr2, "gr3" = gr3)
names(grl) <- c("snparray1", "snparray2", "snparray3")
grl
```

The following data.frame matches masterPheno sample to the GRangesList: 

```{r}
(rangemap <- data.frame(master = c("Jack", "Jill", "Jill"), 
						assay = c("snparray1", "snparray2", "snparray3")))
```

Create an ID map for all available experiments.  Names optional?

```{r}
idmap <- list(exprmap, methylmap, micromap, rangemap)
names(idmap) <- c("expr", "methyl", "micro", "range")
idmap
```

Allowing for the possibility of ID maps entered as dataframes, convert to conventional list: 

```{r}
library(reshape2) # reshape2 used for example only
dfmap <- melt(idmap, id.var = c("master", "assay"))

toListMap(dfmap)
```

Create an named list of experiments `objlist` for the MultiAssay function

```{r}
objlist <- list("Affy" = exprdat, "Methyl 450k" = methyldat, "Mirna" = microdat, "CNV gistic" = grl)
```

Create a toy multiAssayExperiment class object

```{r}
myMultiAssay <- MultiAssay(objlist, masPheno, idmap)
```

Subsetting by Sample

```{r}
logicID <- identifyBySample(myMultiAssay, 1:2)

subMultiAssay <- subsetBySample(myMultiAssay, logicID)

## -------
## TODO: Use Features to Subset myMultiAssay
## -------
## identifyByFeature <- function(MAobject, ){}
```

