
---
title: "dbBacked experiment hub: a demonstration with SQLite"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "Feb 2015"
output:
  BiocStyle::pdf_document:
    toc: yes
    number_sections: yes
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{dbBacked experiment hub: Ovarian cancer TCGA demo}
-->

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

# Introduction

```{r setup,echo=FALSE,results="hide"}
suppressPackageStartupMessages({
library(biocMultiAssay)
library(dplyr)
if (!exists("lovhub2")) example(createHub)
})
```
We have an example "loaded experiment hub" based on
TCGA ovarian tumor data.  It can be created using
```{r lkex,eval=FALSE}
library(biocMultiAssay)
example(createHub)
```
```{r lk2}
lovhub2
```
We don't have a great deal of infrastructure for working with
this.  This document considers a long thin database representation
(not quite triples) of the information.

# Creating and indexing a SQLite representation

A few tools have been coded to facilitate creation of
a SQLite database with the information in `lovhub2`.

```{r lkco}
args(.db_create_long)
args(.db_index)
```

On a modest-size linux machine, these can generate an indexed SQLite
database in a few minutes.

```{r lkco2,eval=FALSE}
newhub = .createOvarianHub()
d1 = .db_create_long("OvarDB.sqlite", newhub, "OvarLong")
d1 = .db_index(d1, "OvarLong")
```

# Some simple queries

Once we have the sqlite generated, we set an
environment variable so that it can be found by R.

```{r sete, eval=FALSE}
Sys.setenv(OVARLONG_PATH="./OvarDB.sqlite")
```
```{r lkinfra}
OvarHub()
OvMasterPheno = function() tbl(OvarHub(), "masterPheno")
OvarLong = function() tbl(OvarHub(), "OvarLong")
tabulateSamples() # will give columns from show(lovhub2)
tabulateFeatures()
```

# A round trip: Creating an ExpressionSet from the database

We use \CRANpkg{tidyr}::spread(), which seems to require data.frame
input at this time.

```{r getes,cache=TRUE}
library(tidyr)
cnv <- OvarLong() %>% filter(assaytype=="OvarCnvGistic")
explain(cnv)
cnvdf = cnv %>% as.data.frame(n=-1)  # not pleasant
cnvByFeat <- cnvdf %>% spread(sampleID, value) 
cnvByFeat[1:5,1:5]
dm = data.matrix(cnvByFeat[,-c(1:3)])
rownames(dm) = cnvByFeat$feature
pd = OvMasterPheno() %>% as.data.frame()
rownames(pd) = pd[,1]
pd = AnnotatedDataFrame( pd[colnames(dm), ] )
OvCnvEs = ExpressionSet( dm, pd )
OvCnvEs
```

# An approach to addressing features in genomic coordinates

We will use the
\Biocpkg{AnnotationDbi} and TxDb infrastructure
in an ad hoc way.  We'll bind gene addresses to the ExpressionSet
instance.

```{r getaddrs}
library(org.Hs.eg.db)
allsym = select(org.Hs.eg.db, keys=keys(org.Hs.eg.db),
   columns="SYMBOL")
mapS2E = allsym$ENTREZID
names(mapS2E) = allsym$SYMBOL
mapS2E = mapS2E[ featureNames(OvCnvEs) ]
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
gnaddrs = genes(TxDb.Hsapiens.UCSC.hg19.knownGene)
mapS2EwAddr = mapS2E[ which(mapS2E %in% names(gnaddrs)) ]
addrd = as.data.frame( gnaddrs[ as.character(mapS2EwAddr) ] )
OvCnvEs = OvCnvEs[ names(mapS2EwAddr), ]
rownames(addrd) = featureNames(OvCnvEs)
fData(OvCnvEs) = addrd
OvCnvEs
fData(OvCnvEs)[1:3,]
```
