---
title: "MultiAssayExperiment: Quick Start Guide"
author: "Marcel Ramos"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Quick-start Guide}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: no
    toc: yes
    toc_depth: 4
---

```{r, echo=FALSE, warning=FALSE}
suppressPackageStartupMessages({
    library(MultiAssayExperiment)
    library(S4Vectors)
    })
```

# Key Features of `MultiAssayExperiment`

## Component slots

### pData - biological units

A `DataFrame` describing the characteristics of the biological units. In
[The Cancer Genome Atlas][] data, for example, the biological units are patients.

Key points: 

* One row per patient
* Zero or more observations in each experiment

```{r}
pheno <- DataFrame(id = 1:4, type = c("a", "a", "b", "b"),
                   sex = c("M", "F", "M", "F"),
                   row.names = c("Bob", "Sandy", "Jake", "Lauren"))
```

### ExperimentList - experiment data

A base `list` or `ExperimentList` object containing the experimental datasets
for the set of samples collected. This gets converted into a class
`ExperimentList` during construction.

Key points:

* Included data classes must support: `[`, `dimnames`, `dim`
* Genomic range-based or ID-based data
* Support open-ended set of data clases

```{r}
dataset1 <- matrix(rnorm(20, 5, 1), ncol = 5,
                  dimnames = list(paste0("GENE", 4:1),
                                  paste0("sample", LETTERS[1:5])))
dataset2 <- matrix(rnorm(12, 3, 2), ncol = 3,
                   dimnames = list(paste0("ENST0000", 1:4),
                                   paste0("samp", letters[1:3])))

expList <- list(exp1 = dataset1, exp2 = dataset2)
expList
```

### sampleMap - relationship graph

A `DataFrame` graph representation of the relationship between the experiments
(`assay` column name), biological units (`primary`), and samples (`colname`).
Helper functions are available for creating a map from a list. See `?listToMap`

Key points:
* relates experimental observations (`colnames`) to `pData`
* permits experiment-specific sample naming, missing, and replicate observations

```{r}
map1 <- DataFrame(primary = c("Bob", "Jake", "Sandy", "Sandy", "Lauren"),
                  colname = paste0("sample", LETTERS[1:5]))
map2 <- DataFrame(primary = c("Jake", "Sandy", "Lauren"),
                  colname = paste0("samp", letters[1:3]))
sampMap <- listToMap(list(exp1 = map1, exp2 = map2))
sampMap
```

<p style="text-align: right;"> <a href="#header">back to top</a> </p>

## MultiAssayExperiment - class constructor function

The `MultiAssayExperiment` constructor function can take three arguments:

1. `experiments` - An `ExperimentList` or `list` of data
2. `pData` - A `DataFrame` describing the biological units
3. `sampleMap` - A `DataFrame` of `assay`, `primary`, and `colname` identifiers

```{r}
(mae <- MultiAssayExperiment(expList, pheno, sampMap))
```

### Subsetting

#### Single bracket `[`

In pseudo code below, the subsetting operations work on the rows of the following indices:
1. _i_ experimental data rows
2. _j_ the primary names or the column names (entered as a `list` or `List`)
3. _k_ assay

```
multiassayexperiment[i = rownames, j = primary or colnames, k = assay]
```

Examples:

```{r}
mae[c("GENE4", "ENST00002"), , ]
mae[, c("Bob", "Jake", "Sandy"), ]
mae[, , "exp1"]
```


#### Double bracket `[[`

The "double bracket" method (`[[`) is a convenience function for extracting
a single element of the `MultiAssayExperiment` `ExperimentList`. It avoids
the use of `experiments(mae)[[1L]]`. For example:

```{r}
mae[[1L]]
```

will extract the first experiment in the `ExperimentList` in the class that
it was stored in. 

<p style="text-align: right;"> <a href="#header">back to top</a> </p>

### Extraction

#### assay and assays

The `assay` and `assays` methods follow `SummarizedExperiment` convention.
The `assay` (singular) method will extract the first element of the
`ExperimentList` and will __return__ a `matrix`. 

```{r}
assay(mae)
```

The `assays` (plurar) method will return a `SimpleList` of the data with each
element being a `matrix`.

```{r}
assays(mae)
```

<p style="text-align: right;"> <a href="#header">back to top</a> </p>

#### Slot accession

Each `slot` in the `MultiAssayExperiment` has its convenient accessor function.
See the table below. 

| Slot | Accessor |
|------|----------|
| `ExperimentList` | `experiments`|
| `pData` | `pData` / `$` * |
| `sampleMap` | `sampleMap` |
| `metadata` | `metadata` |

__*__ The `$` operator on a `MultiAssayExperiment` will return a single
column of `pData`. For example:

```{r}
mae$sex
```

### Transformations 

#### `rearrange`

The `rearrange` function will "reshape" and combine your data into one
`DataFrame`. This is accomplished using either the _long_ or _wide_ format
for the _shape_ argument. 

```{r}
rearrange(mae)
```

For a _wide_ dataset, use the _shape_ argument:

```{r}
rearrange(mae, shape = "wide")[, 1:4]
```

#### `c` - combine

The `c` function allows the user to insert an additional experiment into an
already created `MultiAssayExperiment`.

A `sampleMap` can be provided using in order to map `pData` rows to experiment
column names. In the following example, the "exp3" experiment contains repeated
measurements for Bob.

```{r}
(maec1 <- c(x = mae,
  exp3 = matrix(rnorm(10), ncol = 5,
                dimnames = list(paste0("GENE", c("A", "B")),
                                paste0("sample", LETTERS[1:5]))),
  sampleMap = DataFrame(assay = "exp3",
                        primary = c("Bob", "Bob", "Sandy", "Jake", "Lauren"),
                        colname = paste0("sample", LETTERS[1:5])
                        )
  ))
sampleMap(maec1)
```

For convenience, the _mapFrom_ argument allows the user to map from a particular
experiment **provided** that the **order** of the colnames is in the **same**.
A `warning` will be issued to make the user aware of this assumption.

```{r}
(maec2 <- c(x = mae,
  exp3 = matrix(rnorm(10), ncol = 5,
                dimnames = list(paste0("GENE", c("A", "B")),
                                paste0("sample", LETTERS[1:5]))),
  mapFrom = 1L))
```

## `prepMultiAssay` - Constructor function helper

The `prepMultiAssay` function allows the user to diagnose typical problems
when creating a `MultiAssayExperiment` object. See `?prepMultiAssay` for more
details.

<p style="text-align: right;"> <a href="#header">back to top</a> </p>

# Session info

```{r}
devtools::session_info()
```

<p style="text-align: right;"> <a href="#header">back to top</a> </p>

[The Cancer Genome Atlas]: https://cancergenome.nih.gov/
