---
title: "Using DelayedMatrix with MultiAssayExperiment"
author: "MultiAssay Special Interest Group"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{HDF5Array and MultiAssayExperiment}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
---

```{r, echo=FALSE, warning=FALSE}
suppressPackageStartupMessages({
    library(MultiAssayExperiment)
    library(HDF5Array)
    library(SummarizedExperiment)
})
```

# Integrating an HDF5 backend for MultiAssayExperiment

Dependencies

```{r, eval = FALSE}
library(MultiAssayExperiment)
library(HDF5Array)
library(SummarizedExperiment)
```

## HDF5Array::DelayedArray Constructor

The `HDF5Array` package provides an on-disk representation of large datasets
without the need to load them into memory. Convenient lazy evaluation
operations allow the user to manipulate such large data files based on
metadata. The `DelayedMatrix` class in the `DelayedArray` package provides a
way to connect to a large matrix that is stored on disk.

First, we create a small matrix for constructing the `DelayedMatrix` class.

```{r}
smallMatrix <- matrix(rnorm(10e5), ncol = 20)
```

We add rownames and column names to the matrix object for compatibility with
the `MultiAssayExperiment` representation.

```{r}
rownames(smallMatrix) <- paste0("GENE", seq_len(nrow(smallMatrix)))
colnames(smallMatrix) <- paste0("SampleID", seq_len(ncol(smallMatrix)))
```

Here we use the `DelayedArray` constructor function to create a
`DelayedMatrix` object.

```{r}
smallMatrix <- DelayedArray(smallMatrix)
class(smallMatrix)
head(smallMatrix)
dim(smallMatrix)
```

## Importing HDF5 files

Note that a large matrix from an HDF5 file can also be loaded using the
`HDF5Array` function. 

For example: 

```{r}
dataLocation <- system.file("extdata", "exMatrix.h5", package =
                              "MultiAssayExperiment", mustWork = TRUE)
h5ls(dataLocation)
hdf5Data <- HDF5ArraySeed(file = dataLocation, name = "exMatrix")
newDelayedMatrix <- DelayedArray(hdf5Data)
class(newDelayedMatrix)
head(newDelayedMatrix)
```

### Dimnames from HDF5 file

Currently, the `rhdf5` package does not store `dimnames` in the `h5` file by
default. A request for this feature has been sent to the maintainer of the
`rhdf5` package and any further development to the `HDF5Array` package is
contingent on such lower level dimension name storage.

## Using a `DelayedMatrix` with `MultiAssayExperiment`

A `DelayedMatrix` alone conforms to the `MultiAssayExperiment` API requirements.
Shown below, the `DelayedMatrix` can be put into a named `list` and passed into
the `MultiAssayExperiment` constructor function. 

```{r}
HDF5MAE <- MultiAssayExperiment(experiments = list(smallMatrix = smallMatrix))
sampleMap(HDF5MAE)
colData(HDF5MAE)
```

### `SummarizedExperiment` with `DelayedMatrix` backend

A more information rich `DelayedMatrix` can be created when used in conjunction
with the `SummarizedExperiment` class and it can even include `rowRanges`.
The flexibility of the `MultiAssayExperiment` API supports classes with
minimal requirements. Additionally, this `SummarizedExperiment` with the
`DelayedMatrix` backend can be part of a bigger `MultiAssayExperiment` object.
Below is a minimal example of how this would work:

```{r}
HDF5SE <- SummarizedExperiment(assays = smallMatrix)
assay(HDF5SE)
MultiAssayExperiment(list(HDF5SE = HDF5SE))
```

Additional scenarios are currently in development where an `HDF5Matrix` is
hosted remotely. Many opportunities exist when considering on-disk and off-disk
representations of data with `MultiAssayExperiment`.
