---
title: "Using HDF5Array with MultiAssayExperiment"
author: "MultiAssay Special Interest Group"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{HDF5Array and MultiAssayExperiment}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
---

```{r, echo=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(MultiAssayExperiment))
suppressPackageStartupMessages(library(HDF5Array))
```

# Integrating an HDF5 backend for MultiAssayExperiment

## HDF5Array::DelayedArray Constructor

The `HDF5Array` package provides an on-disk representation of large datasets
without the need to load them into memory. Convenient lazy evaluation
operations allow the user to manipulate such large data files based on
metadata. The `DelayedMatrix` class in the `HDF5Array` package provides a
way to connect to a large matrix that is stored on disk.

First, we create a small matrix for constructing the `DelayedMatrix` class.

```{r}
smallMatrix <- matrix(rnorm(10e5), ncol = 20)
```

We add rownames and column names to the matrix object for compatibility with
the `MultiAssayExperiment` representation.

```{r}
rownames(smallMatrix) <- paste0("GENE", seq_len(nrow(smallMatrix)))
colnames(smallMatrix) <- paste0("SampleID", seq_len(ncol(smallMatrix)))
```

Here we use the `DelayedArray` constructor function to create a
`DelayedMatrix` object.

```{r}
smallMatrix <- DelayedArray(smallMatrix)
class(smallMatrix)
head(smallMatrix)
dim(smallMatrix)
```

## Importing HDF5 files

Note that a large matrix from an HDF5 file can also be loaded using the
`HDF5Dataset` function. 

For example: 

```{r}
dataLocation <- system.file("extdata", "exMatrix.h5", package =
                              "MultiAssayExperiment", mustWork = TRUE)
hdf5Data <- HDF5Dataset(file = dataLocation, name = "exMatrix")
newDelayedMatrix <- DelayedArray(hdf5Data)
class(newDelayedMatrix)
head(newDelayedMatrix)
```
### Dimnames from HDF5 file

Currently, the `rhdf5` package does not store `dimnames` in the `h5` file by
default. A request for this feature has been sent to the maintainer of the
`rhdf5` package.

## Insertting a `DelayedMatrix` into a `MultiAssayExperiment`

```{r}
HDF5MAE <- MultiAssayExperiment(experiments = list(smallMatrix = smallMatrix))
sampleMap(HDF5MAE)
pData(HDF5MAE)
```

